\documentclass{beamer}
\usepackage{lsfolien}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}

\myfootline{System Modelling and Semantic Web -- Winter term 2021}{Hans-Gert
  Gräbe}

\newcommand{\ueberschrift}[1]{\begin{center}\bf #1\end{center}}

\title{Modelling Sustainable Systems\\ and Semantic Web\\[6pt] Information and
  Language \vskip1em}

\subtitle{Lecture in the Module 10-202-2309\\ for Master Computer Science}

\author{Prof. Dr. Hans-Gert Gräbe\\
\url{http://www.informatik.uni-leipzig.de/~graebe}}

\date{December 2021}
\begin{document}

{\setbeamertemplate{footline}{}
\begin{frame}
  \titlepage
\end{frame}}

\section{Technology and Culture}
\begin{frame}{Storytelling and Action}
  
Our actions are closely related to the stories which we are permanently
telling in parallel to each other.
\begin{itemize}
\item With these stories we transcend \emph{our own world of experience} that
  is only a small part of THE WORLD. And we can grasp it only selectively.
\item Storytelling is the form in which we make other people's worlds of
  experience accessible for us.
\item This requires the \emph{coordination of our conceptual worlds}.
\item The \emph{purpose} of the exchange of ideas is to \emph{agree on
  cooperative action}.
\end{itemize}
\end{frame}
\begin{frame}{Storytelling and Action}

The (cooperative) changes of the world are preceded by speaking about these
changes (the imagination of the change).
\begin{itemize}
\item Thinking and Doing: Justified expectations $\to$ World-changing
  cooperative action $\to$ Experienced results
\item Concurrency of cooperate actions.
\item In the tension between justified expectations and experienced results
  the ceasefire lines of the WORLD become visible.
\end{itemize}
\end{frame}

\begin{frame}{Storytelling and Action}
But how change a world that is also constantly changing by itself?
\begin{itemize}
\item Culture: Change the way the world (nature) is changing.
\item Technology (tool perspective) and storytelling (perspective of
  expectations and experiences) are two essential moments of such an
  understanding of culture.
\item Technology comes in here as \emph{general processual knowledge}. 
\item It is only possible to apply it (consciously) \emph{in specific
  contexts}.
\item Separation of goals and means in modern society is a specific form of
  storytelling.
\item The widespread separation of these two moments causes essential problems
  to understand the unity of the reality.
\end{itemize}
\end{frame}

\begin{frame}{Storytelling and Digital Change}
\begin{itemize}
\item Within the digital change new forms of storytelling are developing,
  which break up the previously institutionalised procedures of storytelling.
\item Web 1.0 -- Linked websites as a new form of storytelling.
\item Semantic Web -- RDF as a new basic technology to operate a certain kind
  of storytelling with computer support.
\item Digitisation of important language artifacts. Examples from the German
  Digital Universum:
  \begin{itemize}
  \item German Digital Library (Deutsche Digitale Bibliothek – Kultur und
    Wissen online) \url{https://www.deutsche-digitale-bibliothek.de/} 
  \item Europeana – \url{https://www.europeana.eu/de}
  \item Digital Library at TextGrid –
    \url{https://textgrid.de/digitale-bibliothek}
  \item Zeno.org, a full text library – \url{http://www.zeno.org/}
  \end{itemize}
\end{itemize}\vspace*{2em}
\end{frame}

\begin{frame}{Storytelling and Digital Change}
What social infrastructure conditions are required in order to develop this
potential?
\begin{itemize}
\item Pilar 1: \emph{Free} (as in free speech) \textbf{access to the
  knowledge} resources of the mankind, to communicate prospects of expectation
  and experience in an appropriate way.
\item Pilar 2: \textbf{Acting} in a civil society as responsible
  \emph{private} action, in which the consequences of action are privately
  assigned is a cultural achievement. This includes the ability to close
  contracts, liability, ownership, and institutionalised checks and balances
  in their historical evolution.
\item The digital change requires a new balance between these two pilars. In
  the legal context of a civil society that means above all the readjustment
  of its \emph{legal} constitution and framework.
\end{itemize}
\end{frame}

\section{Information and Language}

\begin{frame}{Information and Language. Linguistics}

  \ueberschrift{What is Language?}
  
It is obviously about processes mediated by language (including computer
language). How does language work?  What does linguistics say about this?
\vskip1em

\begin{quote}\small
  Language, a system of conventional spoken, manual (signed), or written
  symbols by means of which human beings, as members of a social group and
  participants in its culture, express themselves. The functions of language
  include communication, the expression of identity, play, imaginative
  expression, and emotional release.\\
  (\url{https://www.britannica.com/topic/language})
\end{quote}
\end{frame}

\begin{frame}{Information and Language. Linguistics}\small
  
\url{http://de.wikipedia.org/wiki/Sprachsystem}

The idea of how the \textbf{language system} is built depends on which
language or grammar theory it is based. The different theories have mostly the
following assumptions about the components of the language system in common:
\begin{itemize}
\item There are \textbf{linguistic units that are organised hierarchically}
  and reach from the smallest units, the sounds, to the phonemes, morphemes,
  words, parts of sentences, sentences up to texts and possibly to discourses.
\item In this hierarchy, from the morphemes on the units have
  \textbf{additional to their form} a grammatical or lexical \textbf{meaning}.
\item At each level of the hierarchy there are \textbf{rules} that determine
  which positions and combinations of units are allowed and which are not.
  This applies to both the linguistic forms and their meanings.
\end{itemize}
\end{frame}

\begin{frame}{Information and Language. Linguistics}\small
  
\url{http://www.christianlehmann.eu/ling/lg_system/index.html}

\ueberschrift{Formative and Significative Subsystems}

The language system relates thoughts to sounds. This association is indirect
in several ways: A language system cannot associate thoughts ... and also not
sounds ... but only linguistic units with each other. These are on the one
hand \textbf{Significata} (the thought as content of the sign) and on the
other hand \textbf{Significantia} (the sound as expression of the sign).

Hence the language system contains \textbf{two formative subsystems}:
\begin{itemize}
\item In \textbf{Semantics}, the thought is formed into a significatum.
\item In \textbf{Phonology}, the sound is formed into a significant.
\end{itemize}
\end{frame}
\begin{frame}{Information and Language. Linguistics}\small
  
  (cont.)
  
In addition to these formative subsystems, there is the \textbf{Significant
  Subsystem}, which combines Significantia and Significata and thus creates
\textbf{Language Signs}. ... It is divided into two subsystems:
\begin{itemize}
\item Finalised language signs are stored in the \textbf{lexicon}.
\item New language signs are formed in the \textbf{grammar}.
\end{itemize}
\end{frame}

\begin{frame}{Information -- a new Phlogiston?}
\ueberschrift{What is Information?}
\begin{itemize}  
\item Inflationary use of the term information.
  \begin{itemize}
  \item Günter Ropohl remembers the times when there was a counter "Auskunft"
    at a German railway station. (Source: Klemm 2003)
  \end{itemize}
\item The computer scientists stick to an ontologising (and ultimately a
  tangible) concept of information.
\item The linguists talk about language practices.
\end{itemize}
Another critical debate occured in the late 1990s
\begin{itemize}
\item Capurro's Trilemma
\item Trialog (Capurro, Fleissner, Hofkirchner): Is a unified theory of
  information feasible?
\item Heinz Klemm (2003): "A great misery"\\ (German: "Ein großes Elend")
\end{itemize}\bigskip
\end{frame}
\begin{frame}{Information -- a new Phlogiston?}

Peter Janich: The concept of information has necessarily to refer to
successful human communication.\vskip1em
\begin{quote}\small
  However, for successful prompting practices it is fundamental that through
  them a successful connection is established for the involved people between
  the (language) act of prompting and (non-language) act of obeying.  (Janich
  1998)\bigskip

  Grundlegend für gelingende Aufforderungspraxen ist jedoch, dass durch sie
  eine für die beteiligten Personen gelingende Verbindung der (sprachlichen)
  Handlung des Aufforderns und der (gegebenenfalls nicht-sprachlichen)
  Handlung des Befolgens stattfindet.
\end{quote}
\end{frame}

\begin{frame}{Information -- a new Phlogiston?}

Raphael Capurro:\vfill
\begin{quote}\scriptsize
  What I am criticising is the idea to have by the reductionistic concept of
  information \textbf{a kind of phlogiston}: To mean that one comes through
  the different levels -- Aristotle called this logical error \emph{metabis
    eis allo genos} -- and thus to believe e.g. better to explain how life
  arises from matter. So we are not far from the use of the concept of form --
  \emph{informatio} originally goes back to \emph{forma} and \emph{eidos} --
  in relation to matter, life, soul, etc. We would be faced with a new or old
  form of metaphysics.\bigskip

  Was ich kritisiere ist die Vorstellung, durch den reduktionistischen
  Informationsbegriff hätte man eine Art Phlogiston: man kommt durch die
  verschiedenen Ebenen – Aristoteles nannte diesen logischen Fehler metabis
  eis allo genos – und glaubt damit z.B. wie aus Materie Leben entsteht besser
  erklären zu können. Damit sind wir nicht weit von der Verwendung des
  Formbegriffs – informatio geht ursprünglich auf forma und eidos zurück – in
  Bezug auf Materie, Leben, Seele usw. Wir hätten es mit einer neuen oder
  alten Form von Metaphysik zu tun.
\end{quote}
\end{frame}

\begin{frame}{Information -- a new Phlogiston?}

The problem is once again: Where is the human being as an \emph{acting}
subject?

Klaus Fuchs-Kittowski stated already in the 1980s:\bigskip

\begin{quote}\small
  The concept of \emph{unity of self-organisation and generation of
    information} -- the \emph{information processing approach} neglects the
  \emph{formation of meaning} in the process of real life.\bigskip

  Das Konzept der Einheit von Selbstorganisation und Informationsentstehung –
  Der Informationsverarbeitungsansatz vernachlässigt die Bedeutungsbildung im
  Lebensprozess.
\end{quote}\vspace*{1em}
\end{frame}
\end{document}
