\documentclass{beamer}
\usepackage{lsfolien,enumitem}
\usepackage[english]{babel}

\myfootline{System Modelling and Semantic Web -- Spring 2021}{Hans-Gert Gr\"abe}

\newcommand{\ueberschrift}[1]{\begin{center}\bf #1\end{center}}

\title{Modelling Sustainable Systems\\ and Semantic Web\\[6pt] Introduction
  \vskip1em}

\subtitle{Lecture in the Module 10-202-2309\\ for Master Computer Science}

\author{Prof. Dr. Hans-Gert Gräbe\\
\url{http://www.informatik.uni-leipzig.de/~graebe}}

\date{April 2021}
\begin{document}

{\setbeamertemplate{footline}{}
\begin{frame}
  \titlepage
\end{frame}}

\section{Background}
\begin{frame}{Background}

\ueberschrift{Interdisciplinarity}

Guiding motto of the University of Leipzig:
\begin{center}
  A Tradition of Crossing Boundaries
\end{center}

\begin{itemize}
\item[$\bullet$]  Boundaries: Humanities -- Science -- Technology
\item[$\bullet$]  Tradition: The Faculty of Philosophy until 1951
\end{itemize}

But what about technology?

Our mode of production today is closely linked to the use of science and
technology. This brings advantages, but also problems.
\end{frame}

\begin{frame}{Background}
This was not always so. This bourgeois (or capitalist) mode of production has
developed over the last 500 years, first in Western Europe and has since
spread around the globe.

Technology and its use have undergone important metamorphoses during the last
200 years, which had also a decisive influence on the organisational forms of
that mode of production, for example with the establishment of the factory
system and industrial production in the second half of the 19th century.
\end{frame}

\begin{frame}{Background}
This is the beginning of what brought Marx to the vision of a society in which
the "social metabolism" ("gesellschaftlicher Stoffwechsel", MEW 23) is
organised in a way that

"no longer does the worker insert a modified natural thing [Naturgegenstand] as
middle link between the object [Objekt] and himself; rather, he inserts the
process of nature, transformed into an industrial process, as a means between
himself and inorganic nature, mastering it. He steps to the side of the
production process instead of being its chief actor." (MEW 42, ch. 14)
\end{frame}

\begin{frame}{Background}
Marx goes on to state that the development of the productive forces is
\emph{necessarily} heading in such a way towards the organisation of the
"social metabolism".

"But, once adopted into the production process of capital, the means of labour
passes through different metamorphoses, whose culmination is the
\emph{machine}, or rather, an \emph{automatic system of machinery} (system of
machinery: the \emph{automatic} one is merely its most complete, most adequate
form, and alone transforms machinery into a system), set in motion by an
automaton, a moving power that moves itself; this automaton consists of
numerous mechanical and intellectual organs, so that the workers themselves
are cast merely as its conscious linkages." (MEW 42, ch. 13)
\end{frame}

\begin{frame}{Background}
The mastery of that "automaton", that apparently "natural" development of
society, is on the agenda today, because its "naturalness" is increasingly
undermining the very conditions of human existence on our planet.

Marx's vision that in this process the worker "steps to the side of the
production process instead of being its chief actor" is based on a very narrow
understanding of "production process".
\end{frame}

\begin{frame}{Background}
In the 150 years since then this narrow understanding has been replaced by a
common modern understanding of "production process", in which

"it is neither the direct human labour the worker himself performs, nor the
time during which he works, but rather the appropriation of his own general
productive power, his understanding of nature and his mastery over it by
virtue of his presence as a social body – it is, in a word, the development of
the social individual which appears as the great foundation-stone of
production and of wealth". (MEW 42, ch. 14)
\end{frame}

\begin{frame}{Background}
In short, it is a question to overcome the "naturalness" of the "automaton" --
the socio-technical-cultural "apparatus" created by human as social being  --
and put it under the control of the united humanity.

This requires one thing above all -- educated and committed personnel who are
capable of exercising also their civic responsibility (up to Art. 20 of our
constitution).  Academic institutions are requested to deliver an important
contribution to this.  This also requires to cross the old and new boundaries
between Humanities -- Sciences -- Technology.
\end{frame}

\begin{frame}{Background}
The academic education system has been on this path for more than 100 years,
as the development of technological academic educational institutions in
Leipzig shows.
\begin{itemize}
\item[$\bullet$] 1838 Foundation of the Royal Saxonian School of Building
  Professions (K\"oniglich-sächsische Baugewerkeschule) in Leipzig by Albert
  Geutebrück
\item[$\bullet$] 1875 Foundation of the Municipal School of Trades
  (st\"adtische Gewerbeschule) in Leipzig as the historical root for education
  in mechanical and electrical engineering.\medskip

  Realisation that tradesmen need a thorough technical education in addition
  to a general higher education ("humanistische Bildung").
\end{itemize}
\end{frame}

\begin{frame}{Background}
\begin{itemize}
\item[$\bullet$] 1909 Royal Saxonian Building School
\item[$\bullet$] 1914 Technical school for librarians
\item[$\bullet$] 1920 Saxonian State Building School
\item[$\bullet$] 1949 Technical School for Energy Markkleeberg
\item[$\bullet$] 1954 Leipzig College of Civil Engineering
\item[$\bullet$] 1956 Leipzig School of Engineering for Gas Technology
\item[$\bullet$] 1965 School of Engineering for Automation Technology
\item[$\bullet$] 1970 School of Engineering for Energy Management Leipzig
\item[$\bullet$] 1969 Leipzig College of Engineering
\item[$\bullet$] 1977 Unification into Leipzig Technical University
\item[$\bullet$] since 1992 University of Applied Sciences for Technology,
  Economics and Culture
\end{itemize}
\end{frame}

\begin{frame}{Background}
In the 20th century, with \emph{engineers} a whole new professional group
appeared.  Nowadays new professions such as computer scientists, are already
crossing the border between science and technology and can graduate from our
Faculty to obtain a doctorate in science (Dr. rer. nat.), but also a doctorate
in engineering (Dr.-Ing.).  The situation with the Humanities has not yet been
clarified, but at least you can qualify for a Master in Digital Humanities.
\end{frame}

\section{Course Programme}
\begin{frame}{Course Programme}

\ueberschrift{Background and Objectives}

The course is an interdisciplinary offer in the Master in Computer Science and
Master in DH, but can also be taken as a minor (Nebenfach). 

The aim of the course is to illuminate important developments in the outlined
field of development.  
\end{frame}

\begin{frame}{Course Programme}
\ueberschrift{Four Theses}
\begin{itemize}
\item[1)] The short digital age is already over, the corona age started.
\item[2)] Whereas the digital transformation was still characterised by a
  rapidly growing "world of digital data", through the analysis and processing
  of which influence on real-world processes was gained, we are now faced with
  the challenge of using these tools to meet the challenges of the corona
  crisis.
\item[3)] These challenges are only a small foretaste of the challenges that
  climate change will pose.
\item[4)] These challenges are closely linked to fundamental questions not
  only about our economy or mode or production, but also of our understanding
  of technology and science.
\end{itemize}
\end{frame}

\begin{frame}{Course Programme}
It is therefore appropriate to address the three topics

\begin{itemize}
\item[$\bullet$] Social structures of digital change
\item[$\bullet$] Modelling of sustainable systems
\item[$\bullet$] Conceptualisation processes and the Semantic Web
\end{itemize}
and to develop a set of conceptual and terminological tools that are suitable
for a viable analysis of these topics.

The conceptual toolkit to be developed is oriented towards various aspects of
the development of socio-technical systems that are addressed in the lecture
and in the seminar.
\end{frame}

\section{Course Structure}
\begin{frame}{Course Structure}
The course includes

\begin{itemize}
\item[$\bullet$] A lecture "Modelling of Sustainable Systems and Semantic Web"
\item[$\bullet$] A seminar "Complex Systems and Co-Operative Action"
\item[$\bullet$] A TRIZ practical course.
\end{itemize}
\textbf{Note that the access to the e-learning system used in the TRIZ
  practical course is subject to a fee}. Details can be found in the forum of
the OPAL course.
\end{frame}

\section{Organisational Matters}
\begin{frame}{Organisational Matters}
These course parts can be taken for credit in various combinations

\begin{itemize}
\item[1)] All three parts as In-depth Module 10-202-2309 (10 CP) "Modelling
  sustainable systems and semantic web".
  \begin{itemize}[noitemsep]
  \item[$\bullet$] \textbf{Prerequisites for examination:} successfully
    completed seminar and practical course.
  \item[$\bullet$] \textbf{Examination:} oral examination (30 min)
  \end{itemize}
\item[2)] Lecture and seminar as Seminar Module 10-202-2312(5 CP) "Applied
  Computer Science".
  \begin{itemize}[noitemsep]
  \item[$\bullet$] \textbf{Prerequisite for examination:} successfully
    completed seminar.
  \item[$\bullet$] \textbf{Examination:} RDF project and home work paper.
  \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{Organisational Matters}
\begin{itemize}
\item[3)] The practical course alone as Module 10-202-2012 (5 CP) "Current
  Trends in Computer Science".
  \begin{itemize}[noitemsep]
  \item[$\bullet$] \textbf{Prerequisite for examination:} successfully
    completed practical course.
  \item[$\bullet$] \textbf{Examination:} oral examination (30 min)
  \end{itemize}
\end{itemize}
More about this in OPAL \url{https://bildungsportal.sachsen.de/opal} in the
course S21.BIS.SIM.  There, please enrol first in the course and then in the
corresponding group.

You can access OPAL with the data of your studserv account.

\end{frame}

\begin{frame}{Organisational Matters}
You will find a more detailed lecture concept in the github repo
\url{https://github.com/wumm-project/Leipzig-Seminar} in the folder
\texttt{Summerterm-2021}.

\ueberschrift{Data protection}

We follow an Open Culture approach not only theoretically but also practically
and make course materials publicly available. This also applies to the course
materials you have to produce (presentations, handouts) as well as to
(annotated) chat sessions of the seminar discussions, in which your names are
also mentioned. \textbf{We assume your consent to this procedure if you do not
  explicitly object}. The discussions themselves are not recorded.

\end{frame}

\begin{frame}{Organisational Matters}

\begin{itemize}
\item[$\bullet$] Lecture: Thursdays 11:15-12:45, synchronous digital
\item[$\bullet$] The Flipped Classroom Concept
\item[$\bullet$] Continuously updated lecture plan and list of references in
  the \texttt{Lecture/README.md} file in the github Repo.  
\item[$\bullet$] Further (mainly organisational) information also in the forum
  of the OPAL course.
\item[$\bullet$] Seminar: Tuesdays 9:15-10:45, synchronous digital
\item[$\bullet$] All events online in the BBB room BIS.SIM,
  \url{https://meet.uni-leipzig.de/b/gra-w2c-fhz-qnp}
\end{itemize}
\begin{center}\LARGE\bf
  Questions ?
\end{center}

\end{frame}

\end{document}
